---
title: "Getting Started with R and RStudio"
subtitle: "Introduction to R Programming Environment"
description: "This guide helps beginners set up R and RStudio and introduces basic R programming concepts."
---

## Overview

Welcome to the "Getting Started with R and RStudio" guide! This guide is designed to help beginners set up their R programming environment and get familiar with the basics of R and RStudio.

## Installing R and RStudio

1. **Install R**: 

   - Go to the [CRAN website](https://cran.r-project.org/) and download the latest version of R for your operating system (Windows, macOS, or Linux).
   - Follow the installation instructions provided on the website.
2. **Install RStudio**:

   - Visit the [RStudio website](https://rstudio.com/products/rstudio/download/)
   - Download the free version of RStudio Desktop and install it on your computer.
3. **Launch RStudio**: Open RStudio after installation. You should see the RStudio interface with multiple panes.


## RStudio Interface Overview

The RStudio interface consists of several key panes:

- **Source Pane**: This is where you can write and edit your R scripts and R Markdown documents.
- **Console Pane**: This is where you can type and execute R commands directly.
- **Environment/History Pane**: This pane shows the variables and data frames in your current R session, as well as a history of commands you've executed.
- **Files/Plots/Packages/Help/Viewer Pane**: This pane allows you to navigate files, view plots, manage packages, access help documentation, and view web content.

## Writing Your First R Script  

1. In the Source Pane, click on "File" > "New File" > "R Script" to create a new R script.

2. Type the following code into the script:
```{r}
2+2
```

3. To run the code, highlight the line and click the "Run" button or press `Ctrl + Enter` (Windows) or `Cmd + Enter` (Mac). You should see the output `4` in the Console Pane.

## Basic R Commands

Here are some basic R commands to get you started:

- Assigning values to variables:
```{r}
x <- 10
y <- 5
```
- Performing arithmetic operations:
```{r}
sum <- x + y
product <- x * y
```
- Creating a vector:
```{r}
my_vector <- c(1, 2, 3, 4, 5)
```
- Viewing the contents of a variable:
```{r}
print(my_vector)
```

## Installing and Loading Packages

R has a vast ecosystem of packages that extend its functionality. To install and load packages, follow these steps:

1. To install a package, use the `install.packages()` function. For example,
```{r}
#| eval: false
install.packages("ggplot2")
```
2. To load a package into your R session, use the `library()` function:
```{r}
library(ggplot2)
```

## Next Steps

Now that you have set up R and RStudio and learned some basic commands, you can explore more advanced topics in R programming and data analysis. Check out the next guide in this series, [From Basics to Advanced Health Analytics: Exploring Diabetes Data](02-exploring-diabetes-data.qmd), to dive deeper into data analysis using R.


